---
output: rmarkdown::github_document
bibliography: README-refs.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`tstmle`

[![Travis-CI Build Status](https://travis-ci.org/podTockom/tstmle.svg?branch=master)](https://travis-ci.org/podTockom/tstmle)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/podTockom/tstmle?branch=master&svg=true)](https://ci.appveyor.com/project/podTockom/tstmle)
[![Coverage Status](https://img.shields.io/codecov/c/github/podTockom/tstmle/master.svg)](https://codecov.io/github/podTockom/tstmle?branch=master)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)

> Estimation and Inference for Causal Effects with Single Time Series

__Authors:__ Ivana Malenica

## What's `tstmle`?

The `tstmle` package implements targeted maximum likelihood estimation of causal
effects based on the observation of a single time series (i.e., interrupted time
series analysis). The approach considers the case where at each time point in a
time series, a covariate vector, treatment, and outcome are observed in
chronological order. A family of causal effects are defined as the impacts of
stochastic interventions on a subset of the treatment nodes on a future outcome.
This general formulation of the statistical estimation problem subsumes many
other important estimation problems, including but not limited to classical time
series models, group sequential adaptive designs, and even independent and
identically distributed data where the summary measure of the past is simply the
empty set. For details on the theoretical underpinnings of the approach, the
interested reader may consider consulting @vdl2017online. For a general
introduction to the targeted learning methodology and statistical causal
inference, please consult @vdl2011targeted and @vdl2017targeted.

---

## Installation

You can install a stable release of `tstmle` from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/) with:

```{r gh-installation, eval = FALSE}
devtools::install_github("podTockom/tstmle")
```

<!--

In the future, the package will be available from
[CRAN](https://cran.r-project.org/) and can be installed via

```{r cran-installation, eval = FALSE}
install.packages("tstmle")
```

-->

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/podTockom/tstmle/issues).

---

## Example

To illustrate how `tstmle` may be used to ascertain the effect of an
intervention on a single time series, consider the following example:

```{r example}
# Forthcoming...check back later.
```

---

## License

&copy; 2017 [Ivana Malenica](https://github.com/podTockom)

The contents of this repository are distributed under the MIT license. See
below for details:
```
The MIT License (MIT)

Copyright (c) 2017-2018

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.
```

---

## References

