---
title: "Robust Estimation and Inference of Context-Specific Causal Effects based on Observing a Single Time-Series"
author: "Ivana Malenica"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: intro_tstmle.bib
vignette: >
  %\VignetteIndexEntry{Robust Estimation and Inference of Context-Specific Causal Effects based on Observing a Single Time-Series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(warn = FALSE)
```

## Introduction

While studying time-series data, one might be interested what the mean of an outcome at time $\tau$ would have been had we intervened on some of the treatment nodes in the observed time-series. We refer to such a parameter as the marginal parameter, as it reflects the marginal distribution of counterfactual outcomes. We note that the marginal time-series parameter for all binary nodes is implemented in [tstmle01](https://github.com/podTockom/tstmle01/).  

Investigation of the properties of the marginal parameter lead to the conclussion that it cannot be estimated in a double robust manner, however [@c1]. In particular, the asymptotic normality of the marginal paramater relies on a consistent estimation of the whole mechanism of the time-series. Therefore, even for important scenarious where treatment nodes are randomly assigned, the inference still relies on consistent (at rate) estimation of the conditional distributions of the covariate and outcome nodes. 

This raises an important question: are there time-series causal parameters for which we have robust inference when the treatment is known? In light of that, we propose a class of statistical target parameters defined as the average over time $t$ of context-specific pathwise differentiable target parameters of the conditional distribution of the time-series. The `tstmle` package implements several context-specific causal parameters that can be estimated in a double robust manner and therefore fully utilize the sequential randomization. For the interested reader, we refer to [@c2] for the theorems establishing the asymptotic consistency and normality of the targeted maximum likelihood estimator (TMLE) of these causal parameters. 

In particular, `tstmle` implements 3 different context-specific parameters:

1. Average over time of context-specific causal effect of a single time point intervention. 

2. Average over time of context-specific causal effect of multiple time point interventions.

3. Adaptive design learning the optimal individualized rule within a single time-series.

Current implementation supports iterative TMLE, but future releases will support one-step and online TMLE as well [@c3], [@c4]. 

---

## General methodology



---

## Installing and loading the package 

In the following sections, we examine the use of `tstmle` in a variety of simple examples. The package can be installed as follows:

```{r install_pkg, eval=FALSE}
if (!("tstmle" %in% installed.packages())) {
  devtools::install_github("podTockom/tstmle")
}
```

Once the package is installed, we can load it using the following command:

```{r load_pkg, eval=FALSE}
suppressMessages(library(tstmle))
```

---

## Input Data

Input data should be a single time-series of arbitrary size. The longer your provided time-series, better the estimation process, just like in the case of $n$ i.i.d. samples (alas, longer computational demand). The input should be $n$ by $1$ \code{data.frame} with row names indicating which node that particular time point belongs to: $W$, $A$, or $Y$. As mentioned in the previous section, $W$ are the set of covariates, $A$ intervention nodes, and $Y$ outcome. The imposed order must be $A$, $Y$, $W$, so that each time point will be a set of $(A,Y,W)$ for $i = 1, ..., n$. Note that W can contain many different covariate nodes. For more details, see the example simulated dataset in the data directory. 

```{r dataFormat, eval=FALSE}
data("sim_ts_s1")
head(sim_ts_s1)
```

## Context-specific causal effect of a single time point intervention

In this section we examine the effect of a single time-point intervention on a single outcome upstream.
In particular, we intervene on time point $A(3)$ by setting it equal to $1$ with probability $1$, and examine the expected value of $Y(5)$ under this intervened distribution. This simple example is also a special case of a static intervention on node $A(3)$; we could have specified any probability of success here. In order to save time, for all examples considered here we use a shortened time-series located in the data directory. For estimation purposes, we use 100 Monte Carlo draws to estimate conditional means of $Y_{g^*}(\tau_{g^*})$ and 100 Monte Carlo draws to estimate both $h^*$ and $h$-densities. 

```{r SI, eval=FALSE}
#Load relevant packages:
library(sl3)
library(origami)

#Load the data:
data("sim_ts_s1_n50")

#Set library:
Q_library=list("Lrnr_mean", "Lrnr_glm_fast", "Lrnr_glmnet","Lrnr_randomForest","Lrnr_xgboost")
g_library=list("Lrnr_mean", "Lrnr_glm_fast", "Lrnr_glmnet","Lrnr_randomForest","Lrnr_xgboost")

#Obtain estimates:
res<-tstmleSI(sim_ts_s1_n50, Cy=6, Ca=5, V=3)

#TMLE:
res$tmlePsi
#IPTW:
res$iptwPsi

```


## Session Information

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

---

## References

